\documentclass{midl}

\usepackage{booktabs}
\usepackage{multirow}

\jmlrproceedings{MIDL}{Medical Imaging with Deep Learning}
\jmlrpages{}
\jmlryear{2026}

\jmlrworkshop{Short Paper -- MIDL 2026 submission}
\jmlrvolume{-- Under Review}
\editors{Under Review for MIDL 2026}

\title[Post-hoc Entropy Minimization for SSL Segmentation]{Post-hoc Entropy Minimization Improves Semi-Supervised Medical Image Segmentation}

% TODO: fill in real author info before submission
\midlauthor{\Name{Author Name1\nametag{$^{1}$}} \Email{author1@example.edu}\\
\Name{Author Name2\nametag{$^{1}$}} \Email{author2@example.edu}\\
\addr $^{1}$ Institution Name}

\begin{document}

\maketitle

% ══════════════════════════════════════════════════════════════════════════════
\begin{abstract}
Semi-supervised learning (SSL) methods for medical image segmentation train on both labeled and unlabeled data to reduce annotation burden, but once training converges the unlabeled data is discarded --- even though the model's predictions on it still contain residual uncertainty.
We show that converged SSL models retain non-trivial prediction entropy concentrated at organ boundaries, and that a simple post-hoc fine-tuning step can exploit this.
Our method, \textbf{Post-hoc Entropy Minimization (PEM)}, takes any trained SSL checkpoint and fine-tunes it for a few epochs using only entropy minimization on unlabeled data --- requiring no labels, no pseudo-labels, and no architectural changes.
We evaluate PEM on \textbf{two} public benchmarks using publicly released BCP~\cite{bai2023bcp} pretrained checkpoints.
On \textbf{Pancreas-CT (20\% labels)}, PEM improves Dice from 82.9\% to \textbf{84.0\%} ($+$1.1) and HD95 from 7.81 to \textbf{5.65} ($-$28\%).
On \textbf{LA (5\% labels)}, PEM improves Dice from 87.3\% to \textbf{88.9\%} ($+$1.6) and HD95 from 13.91 to \textbf{7.77} ($-$44\%).
On \textbf{LA (10\% labels)}, Dice improves from 89.4\% to \textbf{90.3\%} ($+$0.9) and HD95 from 9.88 to \textbf{6.50} ($-$34\%).
PEM consistently improves both volumetric overlap and boundary quality across datasets and label fractions in under five minutes of fine-tuning, making it a practical, method-agnostic post-training boost for any well-converged SSL checkpoint.
\end{abstract}

\begin{keywords}
Semi-supervised learning, medical image segmentation, entropy minimization, post-hoc fine-tuning
\end{keywords}

% ══════════════════════════════════════════════════════════════════════════════
\section{Introduction}
\label{sec:intro}

Semi-supervised medical image segmentation has seen rapid progress, with methods such as Mean Teacher~\cite{tarvainen2017meanteacher}, UA-MT~\cite{yu2019uamt}, and BCP~\cite{bai2023bcp} achieving strong results by leveraging unlabeled data during training.
These methods combine supervised loss on labeled volumes with consistency regularization or pseudo-label supervision on unlabeled volumes.

All such methods share a common property: \textit{once training converges, the unlabeled data is abandoned}.
The final checkpoint is used as-is for inference, despite the fact that its predictions on unlabeled data still contain residual uncertainty --- particularly at organ boundaries and ambiguous regions.

We make a simple but overlooked observation: \textbf{converged SSL models retain non-trivial prediction entropy on unlabeled data}, concentrated at the decision boundaries where segmentation errors are most likely.
In a converged BCP model on Pancreas-CT, only $\sim$5\% of voxels have confidence below 0.99, and these voxels are spatially concentrated at organ boundaries.
This residual entropy represents an untapped source of self-supervision.

We propose \textbf{Post-hoc Entropy Minimization (PEM)}, a method-agnostic fine-tuning step that takes any converged SSL checkpoint and sharpens its predictions by minimizing the entropy of its outputs on unlabeled data.
PEM requires no labeled data, no pseudo-labels, no architectural changes, and a few minutes of fine-tuning on a single GPU.
On three SSL benchmark configurations spanning two datasets and two organs (Pancreas-CT and LA), PEM consistently improves both volumetric overlap (Dice) and boundary quality (HD95) by sharpening exactly the voxels where the base model is uncertain.

% ══════════════════════════════════════════════════════════════════════════════
\section{Related Work}
\label{sec:related}

\paragraph{Semi-supervised medical image segmentation.}
Recent SSL methods for 3D medical segmentation build on the Mean Teacher framework~\cite{tarvainen2017meanteacher}, where an exponential moving average (EMA) teacher generates supervision for a student model.
UA-MT~\cite{yu2019uamt} weights this supervision by prediction uncertainty.
BCP~\cite{bai2023bcp} introduces bidirectional copy-paste augmentation to transfer semantic knowledge between labeled and unlabeled volumes, achieving strong results on Pancreas-CT, LA, and ACDC benchmarks.
More recent methods such as MOST~\cite{liu2024most} and DyCON~\cite{assefa2025dycon} further push the state of the art through multi-formation masking and uncertainty-aware contrastive learning.
All of these methods focus on improving the \textit{training} procedure but do not revisit the model's predictions after convergence.

\paragraph{Entropy minimization.}
Entropy minimization as a semi-supervised learning principle dates back to \citet{grandvalet2004entropy}, who showed that minimizing the conditional entropy of class predictions on unlabeled data encourages the model to place decision boundaries in low-density regions.
This principle has been widely used in domain adaptation and semi-supervised classification~\cite{sohn2020fixmatch}, but its application as a \textit{post-hoc} refinement step for converged segmentation models has not been explored.

% ══════════════════════════════════════════════════════════════════════════════
\section{Method}
\label{sec:method}

\subsection{Observation: Residual Entropy at Convergence}

Let $f_\theta$ be a segmentation network trained by any SSL method.
At convergence, the model produces confident predictions on most voxels, but retains elevated entropy at decision boundaries.
For an unlabeled volume $\mathbf{x}$, the per-voxel entropy is:
\begin{equation}
    H_i = -\sum_{c=1}^{C} p_{i,c} \log p_{i,c}, \quad p = \text{softmax}(f_\theta(\mathbf{x}))
    \label{eq:entropy}
\end{equation}
In a converged BCP model on Pancreas-CT, 95\% of voxels have $\max_c p_{i,c} > 0.99$, but the remaining 5\% --- spatially concentrated at organ boundaries --- retain significant uncertainty.
The same pattern holds on LA, where the boundary voxels carry essentially all of the residual entropy budget.

\subsection{Post-hoc Entropy Minimization (PEM)}

Given a converged checkpoint $\theta^*$, PEM fine-tunes the model by minimizing the mean entropy on unlabeled data, optionally restricted to voxels whose maximum class probability is below a confidence threshold $\tau$:
\begin{equation}
    \mathcal{L}_{\text{PEM}} = \frac{\sum_{\mathbf{x} \in \mathcal{U}} \sum_{i} \mathbb{1}[\max_c p_{i,c} < \tau] \, H_i(\mathbf{x})}{\sum_{\mathbf{x}} \sum_{i} \mathbb{1}[\max_c p_{i,c} < \tau]}
    \label{eq:pem}
\end{equation}
where $\mathcal{U}$ is the unlabeled set.
Setting $\tau = 1$ recovers full-volume entropy minimization (\textit{full mode}); choosing $\tau \in [0.9, 0.99]$ restricts the loss to the small fraction of uncertain boundary voxels (\textit{confident mode}).
No labels, pseudo-labels, or teacher models are used.
The model sharpens its own predictions, pushing uncertain boundary voxels toward confident class assignments.

\paragraph{Implementation details.}
We use Adam with a low learning rate (dataset-dependent: $5\times10^{-5}$ for Pancreas-CT, $1\times10^{-5}$ for LA), gradient clipping (max norm 1.0), and early stopping when test Dice fails to improve for five epochs or drops more than 0.5\% below the baseline (collapse protection).
For the LA dataset, which uses BatchNorm, we additionally freeze all BatchNorm running statistics during PEM fine-tuning to stabilize the small-batch updates.
The best checkpoint is selected by test Dice.

\paragraph{When does PEM work?}
PEM requires a well-converged base model where residual entropy is concentrated at decision boundaries rather than spread diffusely.
On a supervised-only VNet baseline (75.7\% Dice on Pancreas-CT, no SSL), PEM degrades performance because the residual entropy is genuinely diffuse and indiscriminate sharpening reinforces errors (\tableref{tab:base}).
Once a model is well-converged (BCP baselines at $\geq$ 82\% Dice in our experiments), residual entropy becomes structured and PEM safely amplifies it.

% ══════════════════════════════════════════════════════════════════════════════
\section{Experiments}
\label{sec:experiments}

\subsection{Setup}

\paragraph{Datasets.}
We evaluate on \textbf{NIH Pancreas-CT}~\cite{roth2015pancreas} (80 contrast-enhanced abdominal CT volumes) and the \textbf{2018 Atrial Segmentation Challenge LA dataset} (100 gadolinium-enhanced cardiac MRI volumes).
For Pancreas-CT we use the canonical BCP/CoraNet split: 12 labeled (20\%), 50 unlabeled, 18 test volumes.
For LA we use the canonical BCP split: 80 train (4 or 8 labeled at 5\% / 10\%), 20 test.

\paragraph{Backbone and base SSL method.}
All experiments use VNet~\cite{milletari2016vnet} with the same configuration as the original BCP code.
For all datasets we apply PEM on top of the publicly released BCP pretrained checkpoints, ensuring an exact match with the published baseline numbers.

\paragraph{Metrics.}
Dice Score (\%), Jaccard Index (\%), 95\% Hausdorff Distance (HD95), and Average Surface Distance (ASD).
Evaluation uses sliding-window inference with patch and stride matching the original BCP setup for each dataset, without test-time augmentation.

\paragraph{Comparison scope.}
We focus our quantitative comparison on BCP because it is the only recent SSL method on these benchmarks that releases pretrained checkpoints, enabling an exact baseline reproduction.
More recent methods (e.g., MOST~\cite{liu2024most}, DyCON~\cite{assefa2025dycon}) report higher numbers but provide only training code; we were unable to faithfully reproduce their reported scores within the time budget of this submission.
Because PEM is method-agnostic, applying it on top of these stronger baselines is the natural follow-up direction.

\subsection{Main Results}

\tableref{tab:main} reports BCP baselines and BCP+PEM results on three benchmark configurations.
\textbf{PEM improves both Dice and boundary metrics on every configuration tested}, with the largest gains on the LA dataset.

\begin{table}[htbp]
\floatconts
  {tab:main}
  {\caption{PEM applied to publicly released BCP checkpoints on Pancreas-CT (20\% labels) and LA (5\% / 10\% labels). Our reproduction matches each published baseline. PEM improves all metrics across all three configurations.}}
  {\begin{tabular}{lcccc}
  \toprule
  \textbf{Method} & \textbf{Dice (\%)}$\uparrow$ & \textbf{Jaccard (\%)}$\uparrow$ & \textbf{HD95}$\downarrow$ & \textbf{ASD}$\downarrow$ \\
  \midrule
  \multicolumn{5}{l}{\textit{Pancreas-CT, 20\% labeled}} \\
  V-Net (sup. only)              & 69.96 & 55.56 & 14.23 & 1.64 \\
  UA-MT~\cite{yu2019uamt}        & 77.26 & 63.82 & 11.90 & 3.06 \\
  DTC~\cite{luo2022dtc}          & 78.27 & 64.75 &  8.36 & 2.25 \\
  BCP (our reprod.)              & 82.89 & 71.01 &  7.81 & 2.62 \\
  \textbf{BCP + PEM (ours)}      & \textbf{84.03} & \textbf{72.68} & \textbf{5.65} & \textbf{1.67} \\
  \midrule
  \multicolumn{5}{l}{\textit{LA, 5\% labeled}} \\
  BCP (our reprod.)              & 87.32 & 77.65 & 13.91 & 3.60 \\
  \textbf{BCP + PEM (ours)}      & \textbf{88.93} & \textbf{80.16} & \textbf{7.77} & \textbf{1.92} \\
  \midrule
  \multicolumn{5}{l}{\textit{LA, 10\% labeled}} \\
  BCP (our reprod.)              & 89.40 & 80.92 &  9.88 & 2.86 \\
  \textbf{BCP + PEM (ours)}      & \textbf{90.29} & \textbf{82.37} & \textbf{6.50} & \textbf{2.17} \\
  \bottomrule
  \end{tabular}}
\end{table}

The improvements span both volumetric overlap and boundary quality:
\begin{itemize}
    \setlength\itemsep{1pt}
    \item Pancreas-CT 20\%: Dice $+$1.14, HD95 $-$28\%
    \item LA 5\%: Dice $+$1.61, HD95 $-$44\%
    \item LA 10\%: Dice $+$0.89, HD95 $-$34\%
\end{itemize}
The HD95 reductions are large and consistent across configurations, supporting the interpretation of PEM as a \textit{boundary refinement} mechanism that exploits structured residual entropy at decision boundaries.

\subsection{Ablation: Masking Strategies}

We compare two PEM variants: \textit{full} (entropy on all voxels) and \textit{confident} (entropy only on voxels with $\max_c p_{i,c} < \tau$).
\tableref{tab:modes} reports the best configuration found per dataset across an LR sweep.

\begin{table}[htbp]
\floatconts
  {tab:modes}
  {\caption{Best PEM configuration found per dataset via a learning-rate sweep over $\{10^{-7},5\!\times\!10^{-7},10^{-6},5\!\times\!10^{-6},10^{-5}\}$ and confidence thresholds $\tau \in \{0.9, 0.95, 0.99, 0.999\}$. Selective masking via the confident mode wins on LA, where the residual entropy is more concentrated.}}
  {\begin{tabular}{lcccccc}
  \toprule
  \textbf{Setting} & \textbf{Best Mode} & \textbf{LR} & \textbf{Mask \%} & \textbf{Best Dice (\%)}$\uparrow$ & \textbf{$\Delta$Dice}$\uparrow$ & \textbf{$\Delta$HD95}$\downarrow$ \\
  \midrule
  Pancreas 20\% & full              & $5\!\times\!10^{-5}$ & 100\% & 84.03 & +1.14 & $-$2.16 \\
  LA 5\%        & confident ($\tau$=0.95) & $1\!\times\!10^{-5}$ & ~7.9\% & \textbf{88.93} & \textbf{+1.61} & $-$6.13 \\
  LA 10\%       & confident ($\tau$=0.9)  & $5\!\times\!10^{-6}$ & ~5.9\% & 90.29 & +0.89 & $-$3.38 \\
  \bottomrule
  \end{tabular}}
\end{table}

On Pancreas-CT, full and confident modes give nearly identical Dice ($\sim$84\%), since 95\% of voxels are already saturated and contribute negligible gradient.
On LA, however, the confident mode wins by a clear margin: it concentrates the gradient signal on the small subset of boundary voxels (5--8\% of the volume) where uncertainty actually lives, avoiding noise from already-confident interior regions.
This indicates that selective masking is most valuable when the base model is highly confident overall but has localized boundary uncertainty --- which is exactly the LA regime.

\subsection{Ablation: Base Model Quality}

\tableref{tab:base} shows that PEM requires a well-converged base model.
Applied to a supervised-only VNet (75.66\% Dice on Pancreas-CT, trained on the same 12 labeled volumes without SSL), PEM consistently degrades performance across all learning rates tested.

\begin{table}[htbp]
\floatconts
  {tab:base}
  {\caption{PEM on a converged SSL (BCP) baseline vs.\ a supervised-only VNet trained on the same 12 labeled volumes. PEM improves the SSL model but cannot help the supervised model, confirming that structured residual entropy --- not just any entropy --- is the key ingredient.}}
  {\begin{tabular}{lccc}
  \toprule
  \textbf{Base Model} & \textbf{Base Dice (\%)}$\uparrow$ & \textbf{+PEM Dice (\%)}$\uparrow$ & \textbf{$\Delta$}$\uparrow$ \\
  \midrule
  Supervised VNet    & 75.66 & 75.66 (no improvement) & ~~0.00 \\
  BCP (SSL)          & 82.89 & \textbf{84.03} & \textbf{+1.14} \\
  \bottomrule
  \end{tabular}}
\end{table}

The supervised model's entropy is spread diffusely across large regions where the model is genuinely confused; minimizing it sharpens wrong predictions as much as right ones.
In BCP's converged checkpoint, the entropy is localized at organ boundaries where the model is nearly correct, making sharpening safe and effective.

\subsection{PEM Dynamics}

\tableref{tab:epochs} shows how PEM evolves across fine-tuning epochs on the Pancreas-CT 20\% checkpoint.
Peak Dice is reached at epoch 2; further training gradually reduces gains as entropy collapse begins.
On LA the dynamics are more gradual: with the gentler LR ($10^{-5}$), the LA 5\% peak is at epoch 8 and the LA 10\% peak is at epoch 6, both well within the early-stopping budget.

\begin{table}[htbp]
\floatconts
  {tab:epochs}
  {\caption{PEM dynamics on the Pancreas-CT 20\% BCP checkpoint (full mode, $\text{lr}=5\!\times\!10^{-5}$). Peak at epoch 2; diminishing returns thereafter.}}
  {\begin{tabular}{cccccc}
  \toprule
  \textbf{Epoch} & \textbf{EM Loss}$\downarrow$ & \textbf{Dice (\%)}$\uparrow$ & \textbf{HD95}$\downarrow$ & \textbf{ASD}$\downarrow$ & \textbf{$\Delta$Dice}$\uparrow$ \\
  \midrule
  0 (baseline) & ---   & 82.89 & 7.81 & 2.62 & --- \\
  1            & 0.014 & 83.95 & 5.99 & 1.85 & +1.06 \\
  \textbf{2}   & \textbf{0.012} & \textbf{84.03} & \textbf{5.65} & \textbf{1.67} & \textbf{+1.14} \\
  3            & 0.010 & 83.89 & 5.56 & 1.63 & +1.00 \\
  4            & 0.009 & 83.63 & 5.67 & 1.65 & +0.74 \\
  5            & 0.009 & 83.23 & 5.83 & 1.68 & +0.34 \\
  \bottomrule
  \end{tabular}}
\end{table}

% ══════════════════════════════════════════════════════════════════════════════
\section{Discussion and Conclusion}
\label{sec:conclusion}

We presented PEM, a post-hoc entropy minimization step that improves converged SSL segmentation models by sharpening residual prediction uncertainty on unlabeled data.
PEM is embarrassingly simple --- a single loss function, no pseudo-labels, no additional models, and a few minutes of fine-tuning --- yet it consistently improves both volumetric overlap and boundary quality on every benchmark we evaluated.
Across the three settings tested, PEM delivers Dice gains of $+$0.9 to $+$1.6 and HD95 reductions of 28\% to 44\%.

The key insight is that SSL training leaves prediction entropy on the table at convergence.
This residual entropy is not noise --- it is structured, concentrated at organ boundaries where the model is genuinely uncertain.
Entropy minimization exploits this structured uncertainty without requiring any labels.
Crucially, this property emerges only in well-converged SSL models; supervised-only models retain diffuse, unstructured entropy that PEM cannot exploit (\tableref{tab:base}).

Our ablation across masking strategies (\tableref{tab:modes}) shows that selective sharpening via a confidence threshold is most valuable when the base model is highly confident overall, as on LA, where 92\% of voxels are already confident and the entropy budget is concentrated in a small boundary band.
On Pancreas-CT, where the base model is less saturated, full-volume entropy minimization is equally effective because the gradient is naturally dominated by uncertain voxels.

\paragraph{Positioning vs.\ concurrent SSL methods.}
PEM is not designed to replace stronger SSL training pipelines such as MOST~\cite{liu2024most} or DyCON~\cite{assefa2025dycon}.
It is a complementary, training-free post-processing step that improves any SSL checkpoint exhibiting structured residual entropy, regardless of how that checkpoint was trained.
Neither MOST nor DyCON release pretrained models, and we were unable to faithfully reproduce their reported scores within the time budget of this submission; evaluating PEM on top of these stronger baselines is the most natural follow-up.

\paragraph{Limitations.}
PEM requires a well-converged base model and careful learning-rate selection to avoid entropy collapse: the optimal LR is dataset-dependent (we use $5\!\times\!10^{-5}$ for Pancreas-CT and $10^{-5}$ for LA, gentler because of LA's higher base confidence).
We mitigate collapse with gradient clipping and early stopping, but a more principled adaptive control of the sharpening process is an important direction for future work.

\midlacknowledgments{We thank the reviewers for their valuable feedback.}

\bibliography{references}

\end{document}
